```{r preamble, echo=FALSE, results='hide'}

library(dplyr)
library(devtools)

knitr::opts_chunk$set(warning = FALSE,
                      error = FALSE)

```

# Packing up your data

## Motivation

An R package is the basic unit of reproducible code. There are lots of reasons you might want to make one:

1. As a "personal library" of functions that you use across lots of projects (by copying and pasting).
2. For ctual software that you want to share with others so they can use it, too (things like `lme4`).
3. To hold data from your experiments in a way that's easy to use, easy to share, and reproducibly tracks how the data was processed.

This guide focuses on the third use case. There are lots of good guides for the first case ([Hilary Parker's](https://hilaryparker.com/2014/04/29/writing-an-r-package-from-scratch/) is the classic), and if you're in the second camp you probably already know what you're doing (and if not, Hadley Wickham's [R packages book](http://r-pkgs.had.co.nz/) is an excellent and thorough guide).

We'll start from what used to be my default workflow (raw data file + R scripts to process and make figures etc.), and end up with an R package that allows you to both easily access pre-processed data and tracks how that data was generated from the raw form.

## Preliminaries

0. Install Rstudio.
1. Install `devtools`
2. Configure `devtools`: name, email address
3. Have some data you want to package, plus code to read it into R and process it into a useable form (in a `.R` or `.Rmd` file, etc.)

## What is a package?

To a first approximation, a minimal package is a directory with some R code in `R/` and some metadata in `DESCRIPTION` (what the package is called, who made it, what it does, what other packages it depends on). When you load a package with `library` or `require`, R looks in the package directory and runs the stuff in `R/`.

## Packages can contain data

If there's a `data/` subdirectory in the package, R will also make any data files[^datafiles] there available by name. There are (at least) three ways to load data from a package:

```{r loading-data}
## using namespace
ggplot2::diamonds %>% head()

## using library() (doesn't import into the global environment)
library(ggplot2)
ls()
diamonds %>% head()

## using data() puts it into the global environment
data("diamonds", package="ggplot2") 
ls()
```

[^datafiles]: Data files can be lots of things: `.RData`, `.csv`, `.R`, etc. See `?data`.
